The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Innate nature or tidal nurture?

### 2025-10-20 thesis/abstract.tex
julio final edits
thesis updates

The satellite galaxies of the Milky Way (MW), except the Magellanic
Clouds, are all dwarf spheroidals (dSphs)---gas-free, non-rotating,
dark-matter-dominated stellar systems. The light profiles of dSphs
typically follow an exponential law, with the outer density sharply
declining. Yet systems like the Sculptor (Scl) and Ursa Minor (UMi)
dSphs host member stars out to \(\sim10\) effective radii, indicating
outer deviations from an exponential law. The origin of these extended
profiles is unclear, possibly arising from Galactic tides or intrinsic
properties. In this thesis, we review the evidence for extended light
profiles in Scl and UMi, validating the Bayesian membership catalogue of
\citet{jensen+2024}. To evaluate if tides can produce these density
features, we conduct idealized N-body simulations of both galaxies in
the tidal field of the Milky Way. We find that neither dwarf experiences
tides sufficient to affect their stellar distribution. One complication
is that Sculptor's orbit is strongly influenced by the presence of the
Large Magellanic Cloud (LMC). Our models still suggest that the combined
tides of the LMC and MW are unable to explain Scl's outer profile. We
conclude that the extended light profiles of Scl and UMi are not of
tidal origin. They are instead likely innate, possibly explained by past
mergers, accretion events, or episodic star formation.

>>>>

The satellite galaxies of the Milky Way (MW), except the Magellanic
Clouds, are all dwarf spheroidals (dSphs)---gas-free, non-rotating,
dark-matter-dominated stellar systems. The light profiles of dSphs
typically follow an exponential law, with a sharply declining outer
density. Yet systems like the Sculptor (Scl) and Ursa Minor (UMi) dSphs
host member stars out to \({\sim}10\) effective radii, indicating outer
deviations from an exponential law. The origin of these extended
profiles is unclear, possibly arising from Galactic tides or intrinsic
properties. In this thesis, we review the evidence for extended light
profiles in Scl and UMi, validating the Bayesian membership catalogue of
\citet{jensen+2024}. To evaluate if tides can produce these density
features, we conduct idealized N-body simulations of both galaxies in
the tidal field of the Milky Way. We find that neither dwarf experiences
tides strong enough to affect their stellar distribution. One
complication is that Sculptor's orbit is strongly influenced by the
presence of the Large Magellanic Cloud (LMC). Our N-body models still
suggest that the combined tides of the LMC and MW are unable to explain
Scl's outer profile. We conclude that the extended light profiles of Scl
and UMi are not of tidal origin. They are instead likely innate,
possibly explained by past mergers, accretion events, or episodic star
formation.

### 2025-10-20 thesis/acknowledgements.tex
thesis updates

To all of my new friends I have made along the way. Ruxin, Aviv. To
Pranav, Ted, Jono, Adrianne, Praneet, Will, Maeve, Ewoud. To Akshara for
all your personal and professional advice. We have build a wonderful
community of students in this department. To all the students who have
since moved on and show what we may be: Bobby, Leo, Ryan, Katie,
Fletcher, Ayden, Maheyer, Jess, Dhwani, Spencer, Mallory, Omar, Isaac,
and Dasha.

And to my friends across UVic, Wyatt, Rosemary, Maggie, and the many
people I am still meeting.

To the Linguistics extended family and team \emph{adequate}. Melody,
Lily, Gillian, Stephen, Jamie, Chloë, Alyssa, Kyle, and Marcus. You are
all wonderful human beings, cooks, and always a joy to spend time with.

@@

To my fellow crafters, knitters, and crochet friends. Leise, Wes, Anne,
Anya, Ashley, Sabrina, Sarah. To the graduate student society for
continuing to connect us across departments.

To the international community, and my international friends---Marcela,
Bernardo, Daniel, Walla, Mehrab, and everyone else I have met.

@@

To all of my cousins, but especially Bridget and Gretchen, and my
extended family. And to my siblings for always being a source of joy. To
Arya. To all of my furry friends I have sat---Nano, Pluto, Hobbs,
Penelope, Tochka, Camar, Loki, and Hermes.

>>>>

To all of my new friends I have made along the way. Ruxin, Aviv. To
Pranav, Ted, Jono, Adrianne, Praneet, Will, Maeve, Ewoud, Anne, Anya,
Shoshanah. To Akshara and Simon for all your personal and professional
advice. We have built a wonderful community of students in this
department. To all the students who have since moved on and show what we
may be: Bobby, Leo, Ryan, Katie, Fletcher, Ayden, Maheyer, Jess, Dhwani,
Spencer, Mallory, Omar, Isaac, and Dasha.

And to my friends across UVic, Wyatt, Rosemary, Maggie. To the
Linguistics extended family and team \emph{adequate}. Melody, Lily,
Gillian, Stephen, Jamie, Chloë, Alyssa, Kyle, and Marcus. You are all
wonderful human beings, cooks, and always a joy to spend time with.

@@

To my fellow fibre friends and crafters. Leise, Wes, Ashley, Sabrina,
Sarah. To the graduate student society for continuing to connect us
across departments.

To the international community, and my international friends---Marcela,
Bernardo, Daniel, Victor, Walla, Mehrab, and everyone else I have met.

@@

To Nolan and Olivia, always a joy to see each other. To Bridget and
Gretchen. To my extended family. And to my siblings for always being a
source of joy.

To Arya. To all of my furry friends of friends---Nano, Pluto, Hobbs,
Penelope, Tochka, Camar, Loki, and Hermes.

### 2025-10-20 thesis/data.tex
thesis updates

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/scl_selection.png}
\caption[Sculptor sample selection]{The distributions of various samples
of \emph{Gaia} stars for Sculptor. In each panel, we use light grey
points for the ``all'' sample of stars, turquoise points for ``CMD+PM''
selected stars (Eq.~\ref{eq:sel_cmd_pm}), blue squares for the
``fiducial'' sample (\(P_{\rm sat} > 0.2\)), and indigo diamonds for the
``RV members'' sample. We mark the two far-outlier stars from
\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Left:}
Tangent plane \(\xi, \eta\). The orange ellipses represent 3 and 6
half-light radii. \textbf{Top right:} Colour magnitude diagram in Gaia
\(G\) magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot a
Padova \(12\,\)Gyr, \({\rm [Fe/H] }=-1.68\) isochone in orange. The
black bar in the top left represents the median colour error.
\textbf{Bottom right:} Proper motion in declination \(\mu_\delta\) vs RA
\(\mu_{\alpha*}\) (corrected). The orange point marks the systemic
\citet{MV2020b} proper motion. The black cross represents the median
proper motion error.}\label{fig:scl_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/umi_selection.png}
\caption[Ursa Minor sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline
``velocity confirmed'' members outside a radius of \(6R_h\) with red
stars \citep[from][]{sestito+2023b, pace+2020, spencer+2018}. We also
mark the location of Muñoz 1 with a pink
circle.}\label{fig:umi_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/fornax_selection.png}
\caption[Fornax sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Fornax. The isochrone is instead
for a 2Gyr stellar population. RV measurements are from \citet{WMO2009}.
Fornax does not show the same extended outer halo of probable members as
Sculptor or Ursa Minor.}\label{fig:fornax_selection}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/scl_selection.png}
\caption[Sculptor sample selection]{The distributions of various samples
of \emph{Gaia} stars for Sculptor. In each panel, we use light grey
points for the ``all'' sample of stars, turquoise points for ``CMD+PM''
selected stars (Eq.~\ref{eq:sel_cmd_pm}), blue squares for the
``fiducial'' sample (\(P_{\rm sat} > 0.2\)), and indigo diamonds for the
``RV members'' sample. We mark the two far-outlier stars from
\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Left:}
Tangent plane \(\xi, \eta\). The orange ellipses represent 3 and 6
half-light radii. \textbf{Top right:} Colour magnitude diagram in Gaia
\(G\) magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot a
Padova \(12\,\)Gyr, \({\rm [Fe/H] }=-1.68\) isochone in orange. The
black bar in the top left represents the median member colour error.
\textbf{Bottom right:} Proper motion in declination \(\mu_\delta\) vs RA
\(\mu_{\alpha*}\) (corrected). The orange point marks the systemic
\citet{MV2020b} proper motion. The black cross represents the median
member proper motion error.}\label{fig:scl_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/umi_selection.png}
\caption[Ursa Minor sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline
``velocity confirmed'' members outside a radius of \(6R_h\) with red
stars \citep[from][]{sestito+2023b, pace+2020, spencer+2018}. The
isochrone is 12 Gyr old with \({\rm [Fe/H]} = -2.13\). We also mark the
location of Muñoz 1 with a pink circle.}\label{fig:umi_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/fornax_selection.png}
\caption[Fornax sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Fornax. The isochrone is instead
for a 2Gyr, \({\rm [Fe/H]}=-0.99\) stellar population. RV measurements
are from \citet{WMO2009} and APOGEE. Fornax does not show the same
extended outer halo of probable members as Sculptor or Ursa
Minor.}\label{fig:fornax_selection}
\end{figure}

### 2025-10-20 thesis/extra_results.tex
thesis updates

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}
\caption[Tidal evolution of anisotropy]{The initial and final anisotropy
profiles for the initially anisotropic model of Scl on the \smallperi{}
orbit. The final profile is far more isotropic (\(\beta=0\)) than the
initial.}\label{fig:anisotropy_i_f}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_projected_2d.pdf}}
\caption[Oblate halo projected density snapshots]{The initial and final
(after 9Gyr of tidal evolution) projected density profiles for the
oblate halo on Scl's \smallperi{} orbit, projected on with \(x'\) and
\(z'\) the major and minor axes. The contours are drawn assuming normal
smoothing of 0.4 kpc and are log-spaced with intervals of 0.105
dex.}\label{fig:oblate_i_f}
\end{figure}

@@

Because of the strong tidal interaction with the MW, the trajectory is
substantially perturbed from a point orbit. We adjust the initial
conditions by comparing the change in actions (as calculated in the
static MW-only potential) before and after the pericentre. Our final
model, the \texttt{MW\ impact} model, can approximately reproduce the
observed position of Scl.

@@

>>>>

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}
\caption[Tidal evolution of anisotropy]{The initial and final anisotropy
profiles for the initially anisotropic model of Scl on the \smallperi{}
orbit. The final profile is more isotropic (closer to \(\beta=0\)) than
the initial.}\label{fig:anisotropy_i_f}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_projected_2d.pdf}}
\caption[Oblate halo projected density snapshots]{The initial and final
(after 9Gyr of tidal evolution) projected density profiles for the
oblate halo on Scl's \smallperi{} orbit, projected on with \(x'\) and
\(z'\) the major and minor axes. The contours are drawn assuming normal
smoothing of 0.4 kpc and are log-spaced with intervals of 0.1
dex.}\label{fig:oblate_i_f}
\end{figure}

@@

Because of the strong tidal interaction with the MW, the trajectory is
substantially perturbed from a point orbit. We adjust the initial
conditions by comparing the change in actions (as calculated in the
static MW-only potential) before and after the pericentre. Our final
model, the \texttt{MW\ impact} model, can approximately reproduce the
observed position of Scl (see Fig.~\ref{fig:scl_mw_impact_orbit}). The
model has a Galactocentric initial position of
\([67.83, -352.2, 110.3]\,\kpc\) and velocity of
\([-3.68, 30.79, -22.77]\,\kpc\).

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_impact_orbits.pdf}}
\caption[Sculptor MW impact orbit]{The orbit of Sculptor for the
MW-impact model. The point (dotted) and n-body (solid) diverge by
\(\sim 50\,\kpc\) at early times.}\label{fig:scl_mw_impact_orbit}
\end{figure}

### 2025-10-20 thesis/introduction.tex
thesis updates

Today, a common definition for a dwarf galaxy is that of a
gravitationally bound stellar system with dark matter.\footnote{Or, more
  generally, systems inconsistent with Newtonian dynamics of visible
  matter alone, \citet{willman+strader2012}.} In contrast, star clusters
(like globular clusters) have no clear evidence for dark matter. The
boundary between these two classes blurs for faint, compact stellar
associations. Systems with shared characteristics of globular clusters
and dwarf galaxies, and without evidence as to their nature, are known
as ``ambiguous'' systems \citep[e.g.,][]{smith+2024}.

@@

Most well-studied dwarf galaxies lie in the vicinity of the Milky Way,
the Local Group of galaxies. The Local Group is defined as the group
consisting of galaxies within \(\sim 1\) Mpc from the Milky
Way-Andromeda centre \citep[e.g.,][ and references
therein]{mcconnachie2012}. Today, we know that the Milky Way system is
teeming with dwarfs, many of which are satellites of either the Milky
Way (MW) or Andromeda (M31). Fig.~\ref{fig:mw_satellite_system} shows
the MW satellite system, including dwarf galaxies, globular clusters,
and ambiguous systems. This nearby population of dwarf galaxies is
amenable to resolved, detailed studies aimed at understanding their
nature.

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.png}
\caption[Images of dwarf galaxies]{Images\footnotemark{} of the LMC
\citep[Digitized Sky Survey II,][]{lasker+1996}, Fornax \citep[DES
DR2,][]{abbott+2021}, Sculptor (DES DR2), and Ursa Minor
\citep[UNWISE,][with \textit{Gaia} point sources
overplotted]{lang2014, meisner+lang+schlegel2017, meisner+lang+schlegel2017a}.
The grey ellipse represents the half-light radius for the three dwarf
spheroidals, and the luminosity is derived from the absolute V-band
magnitude of each galaxy.}\label{fig:galaxy_images}
\end{figure}
\footnotetext{Created with hips2fits
  (https://alasky.cds.unistra.fr/hips-image-services/hips2fits), a
  service provided by CDS.}

@@

The very early universe was almost featureless. Our earliest
observations of the universe stem from the cosmic microwave background
(CMB)---revealing a nearly uniform, isotropic blackbody emission. But
tiny perturbations in the CMB, temperature fluctuations of 1 part in
100,000, reveal the underlying seeds of large-scale cosmological
structure. In an expanding universe, gravitational instability makes CDM
overdensities grow and collapse hierarchically onto larger structures.
Initially, baryonic matter was coupled to radiation and resisted
collapse. Dark matter, only influenced by gravity instead, freely
collapsed into the first structures. Mass perturbations sufficiently
small and overdense become self-gravitating structures, known as
\emph{halos}. After recombination, where electrons combined with atomic
nuclei to form atoms, baryons decoupled from radiation and fell into
dark matter halos, where they condensed at the centre through radiative
energy losses. The densest pockets of baryons later formed the first
stars and galaxies.

@@

In \LCDM{} cosmological simulations, dark matter halos are remarkably
self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the authors
observe that the spherically-averaged density profiles \(\rho(r)\) are
universally well described by a two-parameter law,
\begin{equation}\protect\phantomsection\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density. This profile has shown remarkable success at describing \LCDM{}
halos across several orders of magnitude in mass. NFW profiles are
\emph{cuspy}, where the density rises like \(\rho \sim 1/r\) at small
radii \(r \ll r_s\). The steepness of the density profile increases
gradually with radius, and at large radii the density falls off like
\(\rho \sim 1/r^3\). The blue solid curve in Fig.~\ref{fig:nfw_density}
shows an example NFW halo.

@@

\begin{figure}
\centering
\includegraphics[width=3.5in,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
\caption[Example dark matter and stellar density profiles]{Density
profiles in log 3D density versus log 3D radius for stars and dark
matter in a Fornax-like galaxy. The dark matter is more extended and
massive than the star across the entire galaxy. Fornax has a stellar
mass \(M_\star \approx 2.5\times10^7\,\Mo\) with half-light radius
\(0.65\,\kpc\) \citep{munoz+2018, woo+courteau+dekel2008}. The
corresponding cosmological-mean halo for Fornax has \(\vmax=40\,\kpc\)
and \(\rmax=8\,\kpc\), or \(M_{200} = 1\times10^{10}\,\Mo\) and
\(c=12.5\).}\label{fig:nfw_density}
\end{figure}

@@

The right-hand panel of Fig.~\ref{fig:smhm} shows the stellar mass
versus halo mass relation (SMHM, with halo mass represented by
\(\vmax\)) predicted by \LCDM{} cosmological hydrodynamical simulations
of Local Group analogues from the \apostle{} project
\citep{sawala+2016}.\footnote{\apostle{} simulated Local Group analogues
  in a \LCDM{} cosmological context with the hydrodynamical setup from
  the \eagle{} simulations \citep{crain+2015, schaye+2015}}. While there
is some scatter, the range of predicted \(\vmax\) is fairly narrow
across \(\sim 4\) decades in stellar mass. This figure indicates that
the SMHM relation becomes increasingly steep in the dwarf galaxy
regime---many dwarf galaxies are formed in halos of similar mass.
Because lower mass galaxies have shallower potential wells, the
energetic output of evolving stars and possibly supermassive black holes
(i.e., ``feedback'') becomes more effective at removing gas.
Reionization additionally suppresses late star formation in the faintest
galaxies. As a result, the resulting stellar mass of a dwarf galaxy is
highly sensitive to the details of halo assembly and evolution.

@@

Outside the MW system, exponential profiles are common, but sometimes
with modifications. For example, many extragalactic dwarf elliptical,
blue compact, and irregular dwarf galaxies are better described with an
exponential profile to which a central cusp or nuclear region is added
\citep{caldwell+bothun1987, noeske+2003}. On the other hand, some
studies find an inner density decrement relative to exponentials
\citep[e.g.,][]{caldwell+1992, makarov+2012}, or that dwarfs are better
fit by two nested exponentials
\citep[e.g.,][]{aparicio+1997, graham+guzman2003, hunter+elmegreen2006, lee+2018}.
It is unclear how these conclusions apply to the dwarf spheroidals of
the Local Group.

@@

Cosmological simulations struggle to resolve tidal effects on dwarfs.
Since many dwarfs are near the resolution limit, they are vulnerable to
artificial disruption
\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
numerical challenges, idealized simulations model a single subhalo in an
analytic host potential, achieving excellent numerical convergence. For
instance, the simulations we describe later in this thesis reach three
times higher resolution than Aquarius \citep{springel+2008} with 400
times fewer particles. Idealized simulations make numerous
simplifications, neglecting mergers, cosmological context, mass
assembly, and often baryonic physics
\citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
Cosmological simulations appear to predict that tidal disruption is more
common than idealized work on the Milky Way system suggests, but the
role of numerics and assumptions in this discrepancy are unclear
\citep{panithanpaisal+2021, shipp+2023, riley+2024}. We shall use
idealized simulations here to assess tidal effects after the satellite's
infall into the MW halo.

@@

To illustrate the effects of tides on an NFW halo due to a massive host,
we consider a toy model. We evolve an N-body realization of an NFW
subhalo with \(\rmax=5\,\kpc\) and \(\vmax = 27\kms\) orbiting another
NFW static host halo with \(\rmax=25\,\kpc\) and
\(\vmax = 207.4\,\kms\). These choices are motivated by the inferred
structure and masses of dSphs and the Milky Way, respectively. We
simulate the stellar component by assigning stellar weights to each
particle based on the relative densities of the distribution functions
in energy space (see Section~\ref{sec:painting_stars}). The stars
initially follow a 2D exponential profile with a scale radius
\(R_s=0.25\,\kpc\), embedded within the inner dark matter halo. We start
the model at apocentre on an orbit with a pericentre of 20,kpc and
apocentre of 100,kpc. See Section~\ref{sec:methods} for a more detailed
description of our numerical setup.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/idealized_break_radius.pdf}}
\caption[Example tidal simulation]{Example density and velocity
distributions of an idealized dwarf galaxy model undergoing tidal
stripping. \textbf{Top}: The projected 2D stellar density in the
\(y'\)--\(z'\) plane for the initial (left) and final (right)
simulation. \(y'\) is the direction of tangential orbital motion, and
\(z'\) is the direction of orbital angular momentum, as seen from the
centre of the host. The dashed green circle represents the break radius
(Eq.~\ref{eq:r_break}) and the blue arrow points in the orbital
direction. \textbf{Bottom left}: The projected stellar density profile
for the initial (dotted) and final (solid) simulation snapshot.
\textbf{Bottom right}: the mean 3D radial velocity (dot product of
relative position and velocity relative to dwarf centre) as a function
of projected radius. The green arrow marks the break radius in both
lower panels.}\label{fig:idealized_break_radius}
\end{figure}

>>>>

Today, a common definition for a (dwarf) galaxy is that of a
gravitationally bound stellar system with dark matter.\footnote{Or, more
  generally, systems inconsistent with Newtonian dynamics of visible
  matter alone, \citet{willman+strader2012}.} In contrast, star clusters
(like globular clusters) have no clear evidence for dark matter. The
boundary between these two classes blurs for faint, compact stellar
associations. Systems with shared characteristics of globular clusters
and dwarf galaxies, and without evidence as to their nature, are known
as ``ambiguous'' systems \citep[e.g.,][]{smith+2024}.

@@

Most well-studied dwarf galaxies lie in the vicinity of the Milky Way,
the Local Group of galaxies. The Local Group is defined as the group
consisting of galaxies within \(\sim 1\) Mpc from the MW-Andromeda
centre \citep[e.g.,][ and references therein]{mcconnachie2012}. Today,
we know that the Milky Way system is teeming with dwarfs, many of which
are satellites of either the MW or Andromeda (M31).
Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
including dwarf galaxies, globular clusters, and ambiguous systems. This
nearby population of dwarf galaxies is amenable to resolved, detailed
studies aimed at understanding their nature.

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/galaxy_pictures.png}
\caption[Images of dwarf galaxies]{Images\footnotemark{} of the LMC
\citep[Digitized Sky Survey II,][]{lasker+1996}, Fornax \citep[DES
DR2,][]{abbott+2021}, Sculptor (DES DR2), and Ursa Minor
\citep[UNWISE,][with \textit{Gaia} point sources
overplotted]{lang2014, meisner+lang+schlegel2017, meisner+lang+schlegel2017a}.
The grey ellipse represents the half-light radius for the three dwarf
spheroidals, and the luminosity is derived from the absolute V-band
magnitude of each galaxy.}\label{fig:galaxy_images}
\end{figure}
\footnotetext{Created with hips2fits
  (https://alasky.cds.unistra.fr/hips-image-services/hips2fits), a
  service provided by CDS.}

@@

The very early universe was almost featureless. Our earliest
observations of the universe stem from the cosmic microwave background
(CMB)---revealing a nearly uniform, isotropic blackbody emission
\citep[e.g.,][]{ryden2016}. But tiny perturbations in the CMB,
temperature fluctuations of 1 part in 100,000, reveal the underlying
seeds of large-scale cosmological structure. In an expanding universe,
gravitational instability makes CDM overdensities grow and collapse
hierarchically onto larger structures. Initially, baryonic matter was
coupled to radiation and resisted collapse. Dark matter, only influenced
by gravity instead, freely collapsed into the first structures. Mass
perturbations sufficiently small and overdense become self-gravitating
structures, known as \emph{halos} \citep[e.g.,][]{galaxiesbook}. After
recombination, where electrons combined with atomic nuclei to form
atoms, baryons decoupled from radiation and fell into dark matter halos,
where they condensed at the centre through radiative energy losses. The
densest pockets of baryons later formed the first stars and galaxies.

@@

In \LCDM{} cosmological simulations, dark matter halos are remarkably
self-similar. \citet{NFW1996, NFW1997} observe that the
spherically-averaged density profiles \(\rho(r)\) are universally well
described by a two-parameter law,
\begin{equation}\protect\phantomsection\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density. This profile, known by the author's initials NFW, has shown
remarkable success at describing \LCDM{} halos across several orders of
magnitude in mass. NFW profiles are \emph{cuspy}, where the density
rises like \(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness
of the density profile increases gradually with radius, and at large
radii the density falls off like \(\rho \sim 1/r^3\). The blue solid
curve in Fig.~\ref{fig:nfw_density} shows an example NFW halo.

@@

\begin{figure}
\centering
\includegraphics[width=3.5in,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
\caption[Example dark matter and stellar density profiles]{Density
profiles in log 3D density versus log 3D radius for stars and dark
matter in a Fornax-like galaxy. The dark matter is more extended and
massive than the star across the entire galaxy. This galaxy has a
stellar mass \(M_\star \approx 2.5\times10^7\,\Mo\) with half-light
radius \(0.65\,\kpc\) and 2D exponential stars \citep[based
on,][]{munoz+2018, woo+courteau+dekel2008}. The corresponding
cosmological-mean halo has \(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\), or
\(M_{200} = 1\times10^{10}\,\Mo\) and
\(c=12.5\).}\label{fig:nfw_density}
\end{figure}

@@

The right-hand panel of Fig.~\ref{fig:smhm} shows the stellar mass
versus halo mass relation (SMHM, with halo mass represented by
\(\vmax\)) predicted by \LCDM{} cosmological hydrodynamical simulations
of Local Group analogues from the \apostle{} project
\citep{sawala+2016}.\footnote{\apostle{} simulated Local Group analogues
  in a \LCDM{} cosmological context with the hydrodynamical setup from
  the \eagle{} simulations \citep{crain+2015, schaye+2015}.} While there
is some scatter, the range of predicted \(\vmax\) is fairly narrow
across \(\sim 4\) decades in stellar mass. This figure indicates that
the SMHM relation becomes increasingly steep in the dwarf galaxy
regime---many dwarf galaxies are formed in halos of similar mass.
Because lower mass galaxies have shallower potential wells, the
energetic output of evolving stars and possibly supermassive black holes
(i.e., ``feedback'') becomes more effective at removing gas.
Reionization additionally suppresses late star formation in the faintest
galaxies. As a result, the resulting stellar mass of a dwarf galaxy is
highly sensitive to the details of halo assembly and evolution.

@@

Dwarf galaxies outside the Local Group commonly follow exponential
profiles, but sometimes with modifications. For example, many
extragalactic dwarf elliptical, blue compact dwarf, and irregular dwarf
galaxies are better described with an exponential profile to which a
central cusp or nuclear region is added
\citep{caldwell+bothun1987, noeske+2003}. On the other hand, some
studies find an inner density decrement relative to exponentials
\citep[e.g.,][]{caldwell+1992, makarov+2012}, or that dwarfs are better
fit by two nested exponentials
\citep[e.g.,][]{aparicio+1997, graham+guzman2003, hunter+elmegreen2006, lee+2018}.
It is unclear how these conclusions apply to the dwarf spheroidals of
the Local Group.

@@

Cosmological simulations struggle to resolve tidal effects on dwarfs.
Since many dwarfs are near the resolution limit, they are vulnerable to
artificial disruption
\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
numerical challenges, idealized simulations model a single subhalo in an
analytic host potential, achieving excellent numerical convergence. For
instance, the simulations we describe later in this thesis reach three
times higher resolution than Aquarius \citep{springel+2008} with 400
times fewer particles. Idealized simulations make numerous
simplifications, neglecting mergers, cosmological context, mass
assembly, and often baryonic physics
\citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
Cosmological simulations appear to predict that tidal disruption is more
common than idealized simulations of MW dwarfs suggest, but the role of
numerics and assumptions in this discrepancy are unclear
\citep{panithanpaisal+2021, shipp+2023, riley+2024}. We shall use
idealized simulations here to assess tidal effects after the satellite's
infall into the MW halo.

@@

To illustrate the effects of tides on an NFW halo due to a massive host,
we consider a toy model. We evolve an N-body realization of an NFW
subhalo with \(\rmax=5\,\kpc\) and \(\vmax = 27\kms\) orbiting another
NFW static host halo with \(\rmax=25\,\kpc\) and
\(\vmax = 207.4\,\kms\). These choices are motivated by the inferred
structure and masses of dSphs and the Milky Way, respectively. We
simulate the stellar component by assigning stellar weights to each
particle based on the relative densities of the distribution functions
in energy space (see Section~\ref{sec:painting_stars}). The stars
initially follow a 2D exponential profile with a scale radius
\(R_s=0.25\,\kpc\) and total mass \(5\times10^5\,\Mo\), embedded within
the inner dark matter halo. We start the model at apocentre on an orbit
with a pericentre of 20,kpc and apocentre of 100,kpc. See
Section~\ref{sec:methods} for a more detailed description of our
numerical setup.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/idealized_break_radius.pdf}}
\caption[Example tidal simulation]{Example density and velocity
distributions of an idealized dwarf galaxy model undergoing tidal
stripping. \textbf{Top}: The projected 2D stellar density in the
\(y'\)--\(z'\) plane for the initial (left) and final (right)
simulation. \(y'\) is the direction of tangential orbital motion, and
\(z'\) is the direction of orbital angular momentum, as measured from
the centre of the host. The dashed green circle represents the break
radius (Eq.~\ref{eq:r_break}) and the blue arrow points in the orbital
direction. \textbf{Bottom left}: The projected stellar density profile
for the initial (dotted) and final (solid) simulation snapshot.
\textbf{Bottom right}: the mean 3D radial velocity (dot product of
relative position and velocity relative to dwarf centre) as a function
of projected radius. The green arrow marks the break radius in both
lower panels.}\label{fig:idealized_break_radius}
\end{figure}
